---
title: "Homework 3"
output: github_document
---

```{r}
library(tidyverse)
```


## Problem 1
### Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).

Cleaning and combining data sets
```{r}
demographic_df = 
  read_csv("datasets/nhanes_covar.csv",na = c("NA",".",""), skip = 4)|>
  janitor::clean_names() |>
  filter(age >= 21)

accelerometer_df = 
  read_csv("datasets/nhanes_accel.csv",na = c("NA",".",""))|>
  janitor::clean_names() |>
  pivot_longer(
    cols = starts_with("min"),
    names_to = "minute",
    values_to = "activity value"
  )
```

```{r}
cleaned_df = 
  demographic_df |>
  inner_join(accelerometer_df, by=c("seqn"))|>
  filter(!is.na(sex) & !is.na(bmi) & !is.na(education) & !is.na(age)) |>
  mutate(
    sex = ifelse(sex == 1, "male", "female"),
    education = case_when(
      education == 1 ~ "Less than high school",
      education == 2 ~ "High school equivalent",
      education == 3 ~ "More than high school"
  ))
head(cleaned_df)
```


### Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.
```{r}
sex_edu_table = 
  cleaned_df |>
  group_by(sex,education) |>
  mutate(
    education = as.factor(education),
    sex = as.factor(sex)
  ) |>
  summarise(count = n()) |>
  arrange(sex,education)

print(sex_edu_table)
```

```{r}
age_distribution_plot = 
  cleaned_df |>
  ggplot(aes(x = age, fill = sex)) +
  geom_histogram(binwidth = 3) +
  facet_wrap(~education) +
  labs(title = "Age Distribution by Gender and Eduaction", 
       x = "Age",
       y = "Count")
print(age_distribution_plot)
```

comment:

### Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

```{r}
total_activity_df = 
  cleaned_df |>
  group_by(seqn, sex, age, education) |>
  summarise(total_activity = sum(`activity value`, na.rm = TRUE))
```


```{r}
total_activity_df |>
  ggplot(aes(x = age, y = total_activity,color = sex)) +
  geom_point(alpha = .2)+
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~education) +
  labs(title = "Total Activity",
       x = "Age",
       y = "Total Activity",
       color = "Gender") +
  theme_minimal()
```

